Sperm should evolve to make female meiosis fair.

### 2014-05-17 writeup_and_notes/male_dependent_drive.tex
small edits throughout

\author{Yaniv Brandvain \\ email: ybrandvain@gmail.com  \and Graham Coop \\ email: gmcoop@ucdavis.edu }

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0.0pt}
\rfoot{}
\cfoot{\thepage}
\date{}
\bibliographystyle{plain}
\begin{document}
\maketitle
\begin{center}
Center for Population Biology \& Department of Evolution and Ecology \\ University of California - Davis \\ Davis, CA, 95616
\end{center}

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype (Model 1) \gc{REF}. To depict a case
of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm_2_allele_cartoon} and Model 2 in the Supplementary Worksheet)
\footnote{the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.75$? \citep[see ][ for
        discussion]{Buckler1999}. }. %PUT THIS BACK IN
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
\yb{We concluded with an analysis of a related model (Model 3) - in which drivers influence their transmission in females via paternal genotype, rather than sperm haplotype.}

For comparative purposes, we briefly present the standard drive model %as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	%and refer readers to the previous literature on this model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$), a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \yb{\ref{Invasion_space}A}. %, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
\yb{	a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.} 
        % zygotes when it drives. in heterozygotes. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than $(2 d - 1)/(4 d)$. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Invasion_space}).

@@

%OLD YANIV WORDS
% Compared to the invasion criteria, the parameter space
% allowing the fixation of the allele if it is above a cutoff frequency is much broader -- 
% 	a common, recessive, self-promoting driver fixes if
%         $s<(2d-1)/2$ approximately. 
% Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
% 	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
% 	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
% 	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-s_h$)
	further constrains the evolution of a self-promoting driver. 
In fact, with any heterozygous fitness cost, a rare self-promoting
	driver is always selected against. 
However, this case also displays bistability -- 
	when $s$ is sufficiently small (Equation \ref{sForHetFix}) this allele fixes deterministically if its 
	frequency exceeds some threshold,  (Equation \ref{thresh2}, exact results in 
	\yb{FIGURES SXX}).
%	$f_B^*$
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
	observed as stable polymorphisms in natural populations. 
This lack of a balanced polymorphism %(Figure \ref{Invasion_space}) 
	precludes the evolution of an
	allele that suppresses this form of meiotic drive in females.

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
This paternal-genotype dependent model (Model 3) requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and usually fixes when it invades.
	% fixes when $s$ is less than $d_{hom} - 1/2$.
However, when the distorting effect of genotype displays strong dominance in its
	effect on female meiosis ($d_{het}$ is close to $d_{hom}$), 
	there is a a narrow sliver of parameter space sustains a polymorphism
	when the cost of the drive is recessive
	(see Figure S\ref{Effect_of_dominance}, and Equation \ref{polymale}).  
%Presumably this zone of polymorphism comes about because positive frequency 
%	dependence diminishes with the dominance of the male effect driver. 
%As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver (assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
%Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and fixes when $s$ is less than $d_{hom} - 1/2$.
%Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism. 
While mathematically interesting, it does not seem particularly realistic to think that the
	effect of the drive allele would be dominant in its action through the
	male genotype, while the cost would be recessive. 
Therefore, although Model 3 can sustain a polymorphism, 
	the lack of biological reality underlying the small portion of parameter values 
	required for this polymorphism make us doubt it general applicability. 
%Thus we regard this case as interesting but not very general.

@@

\begin{figure}
	\rotatebox{90}{
\includegraphics[width = 0.5
          \textwidth]{Figures/Sperm_egg_cartoon_paper_Fig1.eps}}
\caption{transmission probabilities for alleles through female
  meiosis depend on sperm genotype. 2 allele model}  
\label{Eggsperm_2_allele_cartoon}
\end{figure}

@@

Models 2 and 3, above, explored the dynamics of an allele drove in females when signaled by a complementary signal in sperm.   
We complement this single-locus approach %of an allele with pleiotropic effects in both sexes, 
	with alternative models of two linked loci - one a female driver, 
	and the other, a sperm allele which modifies the effect of drive upon fertilization. 
In this model, a female meiotic driver with no sperm-dependency 
	initially reaches drive-viability equilibrium (with two alleles
	$A$ and $B$ are the ancestral non-driver and driver alleles). 
Subsequently, a sperm-acting modifier of female meiotic drive arises tightly
        linked to this balanced locus (effectively creating a third
        allele/haplotype at this locus). 
Tight linkage offers the best
        chance for a collaboration to evolve between a driver and
       an enhancer of the drive, as recombination breaks up drive haplotypes \citep{Thomson1974,Charlesworth1978,Haig1991}. 
Additionally, tight linkage between female driver and sperm modifier is consistent with the nature of well characterized drive systems which are often maintained as polymorphic inversions with numerous linked modifiers (CITE).

@@

%%%sperm-acting facilitator
We first consider a sperm-acting modifier of
	female meiotic drive tightly linked to the $B$ background (Model 4).
We treat this non-recombining drive/modifier haplotype as a third allele, and we label this allele as $B^+$ when it enhances drive. 
%(i.e. the non-recombining haplotype of a
 %       driver allele and sperm-acting drive modifier)
%The $B^+$ allele 
%	the $B$ and  $B^{+}$ drive in $AB$ and $AB^{+}$ heterozygotes (see Figure \ref{Eggsperm_3_allele_cartoon})
%		generating a third `$B^{+}$' allele (i.e. the non-recombining haplotype of a
   %     driver allele and sperm-acting drive modifier). 
The $B^+$ allele acts in sperm to increase the effectiveness of drive for both
	the $B$ and  $B^{+}$ alleles in $AB$ and $AB^{+}$ heterozygotes (see Figure \ref{Eggsperm_3_allele_cartoon}). 
%	for a simple schematic of this drive model.  
Naively, $B^{+}$ may spread by capitalizing on the additional drive it causes; however,  
	this is not the case for a few simple reasons. 
First, %the sperm-acting drive facilitator (
	the novel $B^{+}$ haplotype
	%) 
	arises when the ancestral driver 
	%($B$) 
	is at drive-selection balance, 
	and therefore immediately suffers a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
	the $B$  allele drive ($B^+$ sperm meeting $AB$ eggs), because $B$ is initially more common than $B^{+}$. 
%As $B^{+}$ facilitates $B$ these interactions frequently form 
%	$BB^{+}$ heterozygotes that suffer the same fitness costs as $BB$. 
Therefore, a sperm-acting drive facilitator alleles experience a profound disadvantage 
	in this scenario, even more so than under the previous two allele model. 
We have found no parameter range of this
	three allele system that allows the sperm-acting drive facilitator $B^{+}$ to
	invade the population \yb{(Mathematical Supplement and Figure XXX)}.

%%%sperm-acting drive suppressors
While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppressors can. 
%	that is sperm-acting alleles that act to restore  fairness of female meiosis, can. %If these sperm-acting drive suppressors arise on
\yb{If a sperm-acting allele that restores fairness to female meiosis arises on
	the drive background (creating a third allele $B^{-}$, Model 4), it displaces the ancestral driver (Equation \ref{coupling}), 
	and often returns to a lower equilibrium frequency than the the $B$ allele, further decreasing the extent of drive in the population. } 
If this sperm-acting drive suppressor arises on 
	the non-driving $A$ background (creating a third allele $A^{-}$, Model 5), 
	or is unlinked to the drive locus (Model 6), it readily invades a population segregating
	for the drive system (see \yb{Figures XXX} and Equations \ref{A-} and \ref{unlinked}).  
This allele lowers the frequency of the original drive system (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}A). 
This result is consistent with previous work showing that drive suppressors unlinked to, or in repulsion phase with drivers usually invade polymorphic drive systems \citep[e.g. ][]{Brandvain2012}.  
\yb{Therefore, all two-locus models of sperm influence on female drive suggest that
	sperm will evolve to oppose female meiotic drive.}

@@

\paragraph{Cost of driver in heterozygotes:} %We explored a more complex model where the self-promoting driver has a cost, $s_h$ in heterozygotes (so $w_{AB} = 1-s_h$) in addition to the cost $s$, in homozygotes.
When the fitness of drive heterozygotes is compromised ($s_h>0$), a self promoting driver cannot invade when rare.
This results from the fact that
when rare %($f_B \ll 1$) 
$B$-bearing sperm rarely
encounter heterozygous eggs ($\sim f_B^2$) but the allele still pays a
cost in heterozygous individuals ($\sim f_B$). 
%Model construction is similar to that above.
%So long as there is any heterozygous fitness cost (i.e. $s_h > 0$) the self-promoting driver cannot invade when rare;
However, this system too, is bistable -- as the driver increases in frequency it is more often fertilized by a driving sperm and therefore drives more effectively. 
Therefore, assuming HWE, if 
\begin{equation}
	s \lessapprox  d -1/2 +s_h(3-2d)/2
	\label{sForHetFix}
\end{equation}
this self-promoting driver deterministically fixes when its frequency is greater than
\begin{equation}
	f_B^* \approx \frac{a_1+\sqrt{a_1^2+8 s_h a_2}}{2 a_2}
	\label{thresh2}
\end{equation}
where, $a_1=1-2 ds_h+4 d s-2 d-3 s_h$ and $a_2=2(s-s_h)(2d-1)$.
Comparison of Equation \ref{thresh2} to exact results obtained by a
simple parameter search (Supplementary Figure \ref{bistable_additive})
show that this approximation is reasonably correct for small
parameter values; \yb{however, this approximation underestimates $f_B^*$ for large parameter values, presumably because they result in strong departures from HWE.}
 %range of other costs to the drive alleles other than a
%simple recessive cost, for example an additive cost. We modified the recursions given in eqn
%\ref{eqn.A1}-\ref{eqn.A3} to feature an additive cost $s$ for each
%copy of the $B$ allele. A none recessive cost to the drive allele can
%cause bistability of the drive allele. The B allele is unable to enter
%the population when rare, this results from the fact that
%when the allele is rare ($x_B \ll 1$) sperm carrying it rarely
%encounter heterozygous eggs ($\sim x^2$) but the allele still pays a
%cost in heterozygous individuals.  By contrast, when the B allele is common ($x_B
%\gg 0$) encounters between heterozygous eggs and $B$ allele sperm
%become more common allowing drive to occur this can allow the $B$
%allele to spread when common. In Supplementary Figure
%\ref{bistable_additive} we show the unstable equilibrium frequency
%above which the $B$ allele can invade. This equilibrium frequency was found by a simple
%parameter search.  \gc{wonder if a bit of work could find this}

@@

@@

\includegraphics[width = 0.5
          \textwidth]{../Scripts/homozyg_bistability.eps}
\caption{Invasion analysis for a self-promoting female meiotic drive allele with
  recessive costs (selection coefficient $s$), showing the region of
  bistability. The colors, and the thin dashed contours, indicate the
  frequency the allele must be above to invade the population (note
  that these alleles reach fixation conditional on invading). In the
  white area the allele can not invade ($x \geq1$), in the solid red
  area the allele can invade and fix when rare. In the left panel we
  show the results obtained by a grid search using the recursion, on
  the right we show the approximation obtained assuming that HWE
  holds. }  
\label{Bistab_homozyg_cost_fig}
\end{figure}

>>>>

\author{Yaniv Brandvain$^1$ \\ email: ybrandvain@gmail.com  \and Graham Coop$^2$ \\ email: gmcoop@ucdavis.edu }

\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{}
\rhead{}
\renewcommand{\headrulewidth}{0.0pt}
\rfoot{}
\cfoot{\thepage}
\date{}
\bibliographystyle{plain}
\begin{document}
\maketitle
\begin{center}
$^1$Department of Plant Biology, \\ University of Minnesota – Twin
Cities, MN, 55108 
$^2$Center for Population Biology \& Department of Evolution and Ecology \\ University of California - Davis \\ Davis, CA, 95616
\end{center}

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype (Model 1) \gc{REF}. The timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.83$ \citep[see ][ for
        discussion]{Buckler1999}.  To depict a case
of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm_2_allele_cartoon} and Model 2 in the Supplementary Worksheet)
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
We concluded with an analysis of a related model (Model 3) - in which
drivers influence their transmission in females via paternal genotype,
rather than sperm haplotype.

For comparative purposes, we briefly present the standard drive model %as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	%and refer readers to the previous literature on this model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare, it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$), a driver cannot fix, and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Invasion_space}A. %, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
        % zygotes when it drives. in heterozygotes. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than $(2 d - 1)/(4 d)$. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Invasion_space}).

@@

%OLD YANIV WORDS
% Compared to the invasion criteria, the parameter space
% allowing the fixation of the allele if it is above a cutoff frequency is much broader -- 
% 	a common, recessive, self-promoting driver fixes if
%         $s<(2d-1)/2$ approximately. 
% Therefore, when $(2 d - 1)/(4 d)<s<(2d-1)/(2)$, 
% 	a recessive, self-promoting driver will deterministically fix if drift, mutation, 
% 	or migration pressure bring its frequency above $(1-2d+4ds)/(2s(2d-1))$, 
% 	but it will be lost otherwise.
Inclusion of a heterozygous fitness cost (i.e. $w_{AB}=1-s_h$)
	further constrains the evolution of a self-promoting driver. 
In fact, with any heterozygous fitness cost, a rare self-promoting
	driver is always selected against. 
However, this case also displays bistability -- 
	when $s$ is sufficiently small (Equation \ref{sForHetFix}) this allele fixes deterministically if its 
	frequency exceeds some threshold,  (Equation \ref{thresh2}, exact results in 
	Figure \ref{bistable_additive}).
%	$f_B^*$
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive})
%While a standard female driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
%	a rare self-promoter pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
%This frequency-independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Below a cutoff frequency the self-promoter driver deterministically decreases in frequency 
%	because it pays the  heterozygous fitness cost but drives
 %      ineffectively (see Figure S\ref{bistable_additive}). 
%Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
%	make drive effective enough to overcome the associated fitness cost.
This bistability prevents self-promoting drivers from invading 	
	reasonably sized populations, and assures that if they do invade, they will rapidly fix.
Our model therefore predicts that self-promoting drivers will not be
	observed as stable polymorphisms in natural populations. 
This lack of a balanced polymorphism %(Figure \ref{Invasion_space}) 
	precludes the evolution of an
	allele that suppresses this form of meiotic drive in females.

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
This paternal-genotype dependent model (Model 3) requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and usually fixes when it invades.
	% fixes when $s$ is less than $d_{hom} - 1/2$.
However, when the distorting effect of genotype displays strong dominance in its
	effect on female meiosis ($d_{het}$ is close to $d_{hom}$), 
	there is a a narrow sliver of parameter space sustains a polymorphism
	when the cost of the drive is recessive
	(see Figure \ref{Effect_of_dominance}, and Equation \ref{polymale}).  
%Presumably this zone of polymorphism comes about because positive frequency 
%	dependence diminishes with the dominance of the male effect driver. 
%As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver (assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
%Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and fixes when $s$ is less than $d_{hom} - 1/2$.
%Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism. 
While mathematically interesting, it does not seem particularly realistic to think that the
	effect of the drive allele would be dominant in its action through the
	male genotype, while the cost would be recessive. 
Therefore, although Model 3 can sustain a polymorphism, 
	the lack of biological reality underlying the small portion of parameter values 
	required for this polymorphism make us doubt it general applicability. 
%Thus we regard this case as interesting but not very general.

@@

\begin{figure}
	\rotatebox{90}{
\includegraphics[width = 0.5
          \textwidth]{Figures/Sperm_egg_cartoon_paper_Fig1.eps}}
\caption{transmission probabilities for alleles through female
  meiosis depend on sperm genotype. 2 allele model. \gc{merge with 2,
    rewrite both captions}}  
\label{Eggsperm_2_allele_cartoon}
\end{figure}

@@

Models 2 and 3, above, explored the dynamics of an allele drove in females when signaled by a complementary signal in sperm.   
We complement this single-locus approach %of an allele with pleiotropic effects in both sexes, 
	with alternative models of two linked loci - one a female driver, 
	and the other, a sperm allele which modifies the effect of drive upon fertilization. 
In this model, a female meiotic driver with no sperm-dependency 
	initially reaches drive-viability equilibrium (with two alleles
	$A$ and $B$ are the ancestral non-driver and driver alleles). 
Subsequently, a sperm-acting modifier of female meiotic drive arises tightly
        linked to this balanced locus (effectively creating a third
        allele/haplotype at this locus). 
Tight linkage offers the best
        chance for a collaboration to evolve between a driver and
       an enhancer of the drive, as recombination breaks up drive haplotypes \citep{Thomson1974,Charlesworth1978,Haig1991}. 
Additionally, tight linkage between female driver and sperm modifier is consistent with the nature of well characterized drive systems which are often maintained as polymorphic inversions with numerous linked modifiers \gc{CITE}.

@@

%%%sperm-acting facilitator
We first consider a sperm-acting modifier of
	female meiotic drive tightly linked to the $B$ background (Model 4).
We treat this non-recombining drive/modifier haplotype as a third allele, and we label this allele as $B^+$ when it enhances drive. 
%(i.e. the non-recombining haplotype of a
 %       driver allele and sperm-acting drive modifier)
%The $B^+$ allele 
%	the $B$ and  $B^{+}$ drive in $AB$ and $AB^{+}$ heterozygotes (see Figure \ref{Eggsperm_3_allele_cartoon})
%		generating a third `$B^{+}$' allele (i.e. the non-recombining haplotype of a
   %     driver allele and sperm-acting drive modifier). 
The $B^+$ allele acts in sperm to increase the effectiveness of drive for both
	the $B$ and  $B^{+}$ alleles in $AB$ and $AB^{+}$ heterozygotes (see Figure \ref{Eggsperm_3_allele_cartoon}). 
%	for a simple schematic of this drive model.  
Naively, $B^{+}$ may spread by capitalizing on the additional drive it causes; however,  
	this is not the case for a few simple reasons. 
First, %the sperm-acting drive facilitator (
	the novel $B^{+}$ haplotype
	%) 
	arises when the ancestral driver 
	%($B$) 
	is at drive-selection balance, 
	and therefore immediately suffers a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
	the $B$  allele drive ($B^+$ sperm meeting $AB$ eggs), because $B$ is initially more common than $B^{+}$. 
%As $B^{+}$ facilitates $B$ these interactions frequently form 
%	$BB^{+}$ heterozygotes that suffer the same fitness costs as $BB$. 
Therefore, a sperm-acting drive facilitator alleles experience a profound disadvantage 
	in this scenario, even more so than under the previous two allele model. 
We have found no parameter range of this
	three allele system that allows the sperm-acting drive facilitator $B^{+}$ to
	invade the population (eqn. \eqref{coupling}, Appendix Model 4, Mathematical Supplement).

%%%sperm-acting drive suppressors
While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppressors can. 
%	that is sperm-acting alleles that act to restore  fairness of female meiosis, can. %If these sperm-acting drive suppressors arise on
If a sperm-acting allele that restores fairness to female meiosis arises on
	the drive background (creating a third allele $B^{-}$, Model 4), it displaces the ancestral driver (Equation \ref{coupling}), 
	and often returns to a lower equilibrium frequency than the the $B$ allele, further decreasing the extent of drive in the population. } 
If this sperm-acting drive suppressor arises on 
	the non-driving $A$ background (creating a third allele $A^{-}$, Model 5), 
	or is unlinked to the drive locus (Model 6), it readily invades a population segregating
	for the drive system (Figure \ref{Eggsperm_3_allele_cartoon} and eqns \ref{A-} and \ref{unlinked}).  
This allele lowers the frequency of the original drive system (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}A). 
This result is consistent with previous work showing that drive suppressors unlinked to, or in repulsion phase with drivers usually invade polymorphic drive systems \citep[e.g. ][]{Brandvain2012}.  
Therefore, all two-locus models of sperm influence on female drive suggest that
	sperm will evolve to oppose female meiotic drive.

@@

\paragraph{Cost of driver in heterozygotes:} %We explored a more complex model where the self-promoting driver has a cost, $s_h$ in heterozygotes (so $w_{AB} = 1-s_h$) in addition to the cost $s$, in homozygotes.
When the fitness of drive heterozygotes is compromised ($s_h>0$), a self promoting driver cannot invade when rare.
This results from the fact that
when rare %($f_B \ll 1$) 
$B$-bearing sperm rarely
encounter heterozygous eggs ($\sim f_B^2$) but the allele still pays a
cost in heterozygous individuals ($\sim f_B$). 
%Model construction is similar to that above.
%So long as there is any heterozygous fitness cost (i.e. $s_h > 0$) the self-promoting driver cannot invade when rare;
However, this system too, is bistable -- as the driver increases in frequency it is more often fertilized by a driving sperm and therefore drives more effectively. 
Therefore, assuming HWE, if 
\begin{equation}
	s \lessapprox  d -1/2 +s_h(3-2d)/2
	\label{sForHetFix}
\end{equation}
this self-promoting driver deterministically fixes when its frequency is greater than
\begin{equation}
	f_B^* \approx \frac{a_1+\sqrt{a_1^2+8 s_h a_2}}{2 a_2}
	\label{thresh2}
\end{equation}
where, $a_1=1-2 ds_h+4 d s-2 d-3 s_h$ and $a_2=2(s-s_h)(2d-1)$.
Comparison of Equation \ref{thresh2} to exact results obtained by a
simple parameter search (Supplementary Figure \ref{bistable_additive})
show that this approximation is reasonably correct for small
parameter values; however, this approximation underestimates $f_B^*$ for large parameter values, presumably because they result in strong departures from HWE.
 %range of other costs to the drive alleles other than a
%simple recessive cost, for example an additive cost. We modified the recursions given in eqn
%\ref{eqn.A1}-\ref{eqn.A3} to feature an additive cost $s$ for each
%copy of the $B$ allele. A none recessive cost to the drive allele can
%cause bistability of the drive allele. The B allele is unable to enter
%the population when rare, this results from the fact that
%when the allele is rare ($x_B \ll 1$) sperm carrying it rarely
%encounter heterozygous eggs ($\sim x^2$) but the allele still pays a
%cost in heterozygous individuals.  By contrast, when the B allele is common ($x_B
%\gg 0$) encounters between heterozygous eggs and $B$ allele sperm
%become more common allowing drive to occur this can allow the $B$
%allele to spread when common. In Supplementary Figure
%\ref{bistable_additive} we show the unstable equilibrium frequency
%above which the $B$ allele can invade. This equilibrium frequency was found by a simple
%parameter search.  \gc{wonder if a bit of work could find this}

@@

\newpage

  \setcounter{figure}{0}\global\def\thefigure{S\arabic{figure}}

@@

\includegraphics[width = \textwidth]{../Scripts/homozyg_bistability.eps}
\caption{Invasion analysis for a self-promoting female meiotic drive allele with
  recessive costs (selection coefficient $s$), showing the region of
  bistability. The colors, and the thin dashed contours, indicate the
  frequency the allele must be above to invade the population (note
  that these alleles reach fixation conditional on invading). In the
  white area the allele can not invade ($x \geq1$), in the solid red
  area the allele can invade and fix when rare. In the left panel we
  show the results obtained by a grid search using the recursion, on
  the right we show the approximation obtained assuming that HWE
  holds. }  
\label{Bistab_homozyg_cost_fig}
\end{figure}
